---
title: "Exploratory Data Analysis"
author: "Hagen Atkeson, Mandy Woo, Allison Dayoan"
date: "5/7/2021"
output:
  pdf_document: default
  html_document: default
---

```{r message = FALSE}
library(tidyverse)
library(magrittr)
library(nlme)
library(mgcv)
library(GGally)

#aids <- read_csv(here::here("data/aids.csv"))
aids <- read_csv("~/Downloads/aids.csv")

# change treatment to factor
aids %<>% 
  mutate(treatment = factor(treatment),
         gender = factor(gender))

glimpse(aids)

# number of participants
num_of_subjects <- n_distinct(aids$id)
num_of_subjects
```
There are 4 relevant covariates: treatment, age, gender, and week. The response variable is log_cd4. There are `r num_of_subjects` subjects.

```{r}
# spaghetti plot
aids %>% 
  ggplot(aes(x = week, y = log_cd4, group = id, color = treatment)) +
  geom_point() + 
  geom_line()
```
The data is unbalanced, measurements were not taken at the same occasions. Therefore, Analysis of Response Profiles cannot be used.

### Univariate Summaries
```{r}
summary(aids)
```
The dataset is overwhelmingly male, in accordance with how males make up the majority of those suffering from AIDS. The measurement occasions span 40 weeks, with the mean being being less than 20, indicating that there are more observations in the first half of the study than the second half. Although counting distinct ids give a figure of `r num_of_subjects`, we see instead that the max id is 1313, meaning that some ids may be missing.


Some univariate graphs depicting the distribution of ages, weeks, and genders studied:
```{r}
aids_wide <- spread(aids, week, log_cd4)

aids_wide %>% 
  ggplot(aes(x = age)) +
  geom_histogram() +
  labs(x = "Age (Years)") +
  ggtitle("Count of Ages")

aids %>% 
  ggplot(aes(x = week)) +
  geom_histogram() +
  labs(x = "Time Since Baseline (weeks)") +
  ggtitle("# of Participations During Each Week")

aids_wide %>% 
  ggplot(aes(x = gender)) +
  geom_bar() +
  labs(x = "Gender") +
  ggtitle("# Participants by Gender")
```
The univariate summaries support that there is imbalances in participants regarding gender. There are more than 7 times the number of males than females in this study. The majority of the participants are between 30 and 40. The times when measurements are taken varies a lot.


### Bivariate Summaries

Relationship between covariates:
```{r}
ggpairs(aids, columns = c("treatment", "age", "week", "gender", "log_cd4"))
```

CD4 counts over time, grouped by gender:
```{r}
aids <- aids %>% 
  group_by(id) %>% 
  mutate(occasion = row_number())

aids %>% 
  group_by(treatment, gender, occasion) %>% 
  summarize(mean_cd4 = mean(log_cd4)) %>% 
  ggplot(aes(x = occasion, y = mean_cd4, color = factor(treatment))) +
  geom_point() +
  geom_line() +
  labs(x = "Occasion", y = "Mean Log Transformed CD4 Counts (cells/mm^3)", color = "Treatment Group") +
  facet_wrap(~ gender) +
  ggtitle("Mean Log Transformed CD4 Counts By Occasion Differentiated by Treatment and Gender")
```
Treatment group 4 significantly decreases female CD4 count, while the same can be said about treatment group 2 for males.  Overall, Treatment group 3 increases average CD4 counts most for females, and all treatments decrease average CD4 counts for males.


CD4 counts over time, grouped by age:
```{r}
aids <- aids %>% 
  mutate(age_group = case_when(age > 70 ~ "70+",
                               70 >= age & age > 60 ~ "60-70", 
                               60 >= age & age > 50 ~ "50-60", 
                               50 >= age & age > 40 ~ "40-50", 
                               40 >= age & age > 30 ~ "30-40", 
                               30 >= age & age > 20 ~ "20-30", 
                               20 >= age ~ "less than 20"))

aids %>% 
  group_by(treatment, age_group, occasion) %>% 
  summarize(mean_cd4 = mean(log_cd4)) %>% 
  ggplot(aes(x = occasion, y = mean_cd4, color = age_group)) +
  geom_point() +
  geom_line()+
  labs(x = "Occasion", y = "Mean Log Transformed CD4 Counts (cells/mm^3)", color = "Age") +
  facet_wrap(~ treatment) +
  ggtitle("Mean Log Transformed CD4 Counts By Occasion Differentiated by Treatment and Age")
```
Treatments 3 and 4 increase the average log CD4 count for the 60-70 age group.  Treatment 3 works best for the 20-30 age group.  For the 30-40 and 40-50 age group, all the treatments decreased average log CD4 count.  For the < 20 age group, only the 2nd treatment group increased average log CD4 count, but there are not many individuals in this age group studied.

Mean log_cd4 by week
```{r}
log_cd4_mean <- aids %>% group_by(treatment, week) %>%
  summarize(mean=mean(log_cd4, na.rm = TRUE))
ggplot(log_cd4_mean, aes(x = week, y = mean, group=treatment, color=as.factor(treatment))) +
  geom_line() +
  facet_wrap(~ treatment) +
  labs(x='Week', y = 'Mean Log of CD4', color='Treatment') +
  ggtitle("Mean Log Transformed CD4 Counts By Week Differentiated by Treatment")
```
The mean of log cd4 of treatment1 is rarely increases over 4, but otherwise, there is not a big distinction between the mean log cd4.

```{r}
aids %>% 
  group_by(treatment, occasion) %>% 
  summarize(mean_cd4 = mean(log_cd4)) %>% 
  ggplot(aes(x = occasion, y = mean_cd4)) +
  geom_point() +
  geom_line()+
  labs(x = "Occasion", y = "Mean Log Transformed CD4 Counts (cells/mm^3)") +
  facet_wrap(~ treatment) +
  ggtitle("Mean Log Transformed CD4 Counts By Occasion Differentiated by Treatment")
```
The mean log cd4 tend to decrease after each measurement.

### Outliers

```{r}
# age
aids_wide %>% 
  ggplot(aes(x = age)) +
  geom_histogram()
```
The data with regards to age is right-skewed, but approximately normal. There are no apparent outliers.

```{r}
# number of observations per subject
aids %>% 
  group_by(id) %>% 
  count() %>%
  mutate(n = factor(n, levels = c("1", "2", "3", "4", "5", "6", "7", "8", "9"))) %>% 
  ggplot(aes(x = n)) + 
  geom_bar() + 
  labs(x = "Observations per individual")

over_6 <- aids %>% 
  group_by(id) %>% 
  count() %>%
  filter(n > 6) %>% 
  nrow()
```
In terms of how many measurements were taken for each individual, there are `r over_6` individuals who have more than 6 measurements.
