---
title: "Exploratory Data Analysis"
author: "Hagen Atkeson, Mandy Woo"
date: "5/7/2021"
output: html_document
---

```{r message = FALSE}
library(tidyverse)
library(magrittr)

aids <- read_csv(here::here("data/aids.csv"))

# change treatment to factor
aids %<>% 
  mutate(treatment = factor(treatment),
         gender = factor(gender))

glimpse(aids)

# number of participants
num_of_subjects <- n_distinct(aids$id)
```
There are 4 relevant covariates: treatment, age, gender, and week. The response variable is log_cd4. There are `r num_of_subjects` subjects.

```{r}
# spaghetti plot
aids %>% 
  ggplot(aes(x = week, y = log_cd4, group = id, color = treatment)) +
  geom_point() +
  geom_line()
```
The data is unbalanced, measurements were not taken at the same occasions. Therefore, Analysis of Response Profiles cannot be used.

### Univariate Summaries
```{r}
summary(aids)
```
The dataset is overwhelmingly male, in accordance with how males make up the majority of those suffering from AIDS. The measurement occasions span 40 weeks, with the mean being being less than 20, indicating that there are more observations in the first half of the study than the second half. Although counting distinct ids give a figure of `r num_of_subjects`

### Bivariate Summaries
```{r}

```


### Outliers

```{r}

```

