---
title: "Final Report - First Draft"
author: "Hagen Atkeson, Allison Dayoan, Mandy Woo"
date: "6/3/2021"
output:
     pdf_document:
         latex_engine: xelatex
---
```{r include = FALSE}
knitr::opts_chunk$set(echo=FALSE)
knitr::opts_chunk$set(message=FALSE)
```

## Exploratory Data Analysis

```{r message = FALSE, echo = FALSE}
library(tidyverse)
library(magrittr)
library(nlme)
library(mgcv)
library(GGally)
library(lme4)

#aids <- read_csv(here::here("data/aids.csv"))
aids <- read_csv("~/Downloads/aids.csv")

# change treatment to factor
aids %<>% 
  mutate(treatment = factor(treatment),
         gender = factor(gender))

# add a column for occasion
aids <- aids %>% 
  group_by(id) %>% 
  mutate(occasion = row_number())

glimpse(aids)

# number of participants
num_of_subjects <- n_distinct(aids$id)
```
There are 4 relevant covariates: treatment, age, gender, and week. The response variable is log_cd4. There are `r num_of_subjects` subjects.

```{r}
# spaghetti plot
aids %>% 
  ggplot(aes(x = week, y = log_cd4, group = id, color = treatment)) +
  geom_point() + 
  theme_light() +
  labs(title = "Log(cd4) by Week")
  
# mean response time plot
aids %>% 
  group_by(treatment, occasion) %>% 
  summarize(mean_log_cd4 = mean(log_cd4)) %>% 
  ggplot(aes(x = occasion, y = mean_log_cd4, group = treatment, color = treatment)) +
  geom_point() +
  geom_line() +
  labs(x = "Occasion", y = "Mean Log(cd4)") +
  ggtitle("Mean Response Time Plot")
```
Looking at the data by individual, there doesn't seem to be any significant outliers. The data is unbalanced, measurements were not taken at the same occasions. Therefore, Analysis of Response Profiles cannot be used.

### Univariate Summaries
```{r}
summary(aids) %>% 
  knitr::kable()

aids_wide <- spread(aids, week, log_cd4)

aids_wide %>% 
  ggplot(aes(x = age)) +
  geom_histogram() +
  labs(x = "Age (Years)") +
  ggtitle("Count of Ages") +
  theme_light()

aids %>% 
  ggplot(aes(x = week)) +
  geom_histogram() +
  labs(x = "Time Since Baseline (weeks)") +
  ggtitle("# of Participations During Each Week") +
  theme_light()

aids_wide %>% 
  ggplot(aes(x = gender)) +
  geom_bar() +
  labs(x = "Gender") +
  ggtitle("# Participants by Gender") +
  theme_light()
```
The dataset has nearly eight times more males than females, in accordance with how males make up the majority of those suffering from AIDS. In terms of age, the majority of the participants are between 30 and 40, and the data is skewed right.  
The measurement occasions span 40 weeks, with the mean being being less than 20, indicating that there are more observations in the first half of the study than the second half. The times when measurements are taken varies a lot.

## Bivariate Summaries
CD4 counts over time, grouped by gender:
```{r}
aids %>% 
  group_by(treatment, gender, occasion) %>% 
  summarize(mean_cd4 = mean(log_cd4)) %>% 
  ggplot(aes(x = occasion, y = mean_cd4, color = factor(treatment))) +
  geom_point() +
  geom_line() +
  labs(x = "Occasion", y = "Mean Log Transformed CD4 Counts (cells/mm^3)", color = "Treatment Group") +
  facet_wrap(~ gender) +
  ggtitle("Mean Log Transformed CD4 Counts By Occasion Differentiated by Treatment and Gender")
```
Treatment group 4 significantly decreases female CD4 count, while the same can be said about treatment group 2 for males.  Overall, Treatment group 3 increases average CD4 counts most for females, and all treatments decrease average CD4 counts for males.  This discrepancy indicates that a gender/treatment interaction term may be good to include in our model.

CD4 counts over time, grouped by age:
```{r}
aids <- aids %>% 
  mutate(age_group = case_when(age > 70 ~ "70+",
                               70 >= age & age > 60 ~ "60-70", 
                               60 >= age & age > 50 ~ "50-60", 
                               50 >= age & age > 40 ~ "40-50", 
                               40 >= age & age > 30 ~ "30-40", 
                               30 >= age & age > 20 ~ "20-30", 
                               20 >= age ~ "less than 20"))
aids %>% 
  group_by(treatment, age_group, occasion) %>% 
  summarize(mean_cd4 = mean(log_cd4)) %>% 
  ggplot(aes(x = occasion, y = mean_cd4, color = age_group)) +
  geom_point() +
  geom_line()+
  labs(x = "Occasion", y = "Mean Log Transformed CD4 Counts (cells/mm^3)", color = "Age") +
  facet_wrap(~ treatment) +
  ggtitle("Mean Log Transformed CD4 Counts By Occasion Differentiated by Treatment and Age")
```
Treatments 3 and 4 increase the average log CD4 count for the 60-70 age group.  Treatment 3 works best for the 20-30 age group.  For the 30-40 and 40-50 age group, all the treatments decreased average log CD4 count.  For the < 20 age group, only the 2nd treatment group increased average log CD4 count, but there are not many individuals in this age group studied.  The difference in treatment effects for each age group shows that an age/treatment interaction term may be beneficial for the model.

## LME Model
As the timing and number of measurement occasions differ by individual, a linear mixed-effects model would be more suited to the data than a general least squares model. Here is the linear mixed effects model:  
$E(log\_cd4_{ij})=\beta_1week_{ij} + \beta_2treatment_{i}\times week_{ij}+ \beta_3gender_i\times treatment_i+ \beta_4age_i\times treatment_i$  
According to the exploratory data analysis, the interaction between treatment and gender, and treatment and age are potentially interesting. Here we shall test if they, along with the interaction term between treatment and week, are needed:  
$H_0:\beta_2=\beta_3=\beta_4=0$  
$H_1:H_0$ is not true. 
```{r}
aids %<>% 
  mutate(treatment = factor(treatment, levels = c(1, 2, 3, 4)),
         gender = factor(gender, levels = c("male", "female")))

model_rand_interaction <- lme(log_cd4 ~ week + treatment:week + gender:treatment + age:treatment,
                              data = aids,
                              random = ~ week | id,
                              method = "ML")

model_rand <- lme(log_cd4 ~ week,
                  data = aids,
                  random = ~ week | id,
                  method = "ML")

summary(model_rand_interaction)
summary(model_rand)

anova(model_rand, model_rand_interaction)
```
Here, we use the ANOVA test to compare the two models and their log-likelihoods.    
$With a p-value of <.0001, there is statistical evidence to reject the null hypothesis that \beta_2=\beta_3=\beta_4=0.$
Therefore, the model with the interaction terms is better.  


## GLME Model
```{r}
aids %<>% 
  mutate(cd4 = round(exp(log_cd4) - 1),
         week_sqr = week^2)

model_glmer <- glmer(cd4 ~ week + week_sqr + treatment:week + gender:treatment + age:treatment + (1 + week | id),
               family = poisson,
               nAGQ = 0,
               data = aids)

model_lme <- lme(log_cd4 ~ week + week_sqr + treatment:week + gender:treatment + age:treatment,
                              data = aids,
                              random = ~ week | id,
                              method = "ML")

summary(model_glmer)
summary(model_lme)
```
When comparing AIC, the model produced by the `lme()` function has the lower value of 11998.15, compared to the model produced by `glmer()`, which is 52897.8. The lme model is therefore superior.

## Residual Analysis
```{r}
res_population <- residuals(model_lme, type = "response", level = 0)
Sigma_i <- extract.lme.cov(model_lme, aids)
L_i <- t(chol(Sigma_i)) 
res_transformed <- solve(L_i) %*% res_population

# histogram
tibble(r_star = res_transformed) %>% 
  ggplot(aes(x = r_star)) +
  geom_histogram(aes(y = stat(density)), bins = 14, color = "black", fill = "gray") +
  geom_function(fun = dnorm, color = "blue") +
  labs(x = "Residuals", y = "Density") +
  ggtitle("Histogram of Transformed Residuals")

# QQPlot
tibble(r_star = res_transformed) %>% 
  ggplot(aes(sample = r_star)) +
  geom_qq_line(color = "blue") +
  geom_qq(shape = 1) +
  labs(x = "Quantiles of Standard Normal", y = "Quantiles of Transformed Residuals") +
  ggtitle("QQ Plot")
```
The QQPlot shows us that the transformed residuals have heavier tails than what would be expected for data from the Normal distribution, and that perhaps the assumption of normality is invalid.

```{r}
# scatter plot of predicted value vs. residuals
mu_hat <- fitted(model_lme, level = 0)
mu_hat_transformed <- solve(L_i) %*% mu_hat
tibble(x = mu_hat_transformed, y = res_transformed) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Transformed Predicted Value", y = "Transformed Residual") +
  ggtitle("Predicted Values vs. Transformed Residuals")

# scatter plot of time vs. residuals
time <- aids$week
time_transformed <- solve(L_i) %*% time
tibble(x = time_transformed, y = res_transformed) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Transformed Time", y = "Transformed Residual") +
  ggtitle("Time vs. Transformed Residuals")
```
Both scatterplots show that the mean of the transformed residuals stays approximately zero in regard to the predicted value and over time. This indicates that the assumed model of variance induced by the linear mixed-effects is adequate.

```{r}
# Semi-variogram
Variogram(model_lme,
          data = aids,
          form = ~ week | id,
          resType = "normalized") %>% 
  as_tibble() %>% 
  ggplot(aes(x = dist, y = variog)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.3) +
  ggtitle("Semi-Variogram")
```
The semi-variogram has randomly distributed points around the loess line, so there is no trend over time.  This indicates the model is good.

```{r}
# Mahalanobis Distance
mahalanobis_data <- tibble(id = aids$id, r_star = res_transformed) %>% 
  group_by(id) %>% 
  nest()
mahalanobis_data <- mahalanobis_data %>% 
  mutate(df = map_dbl(data, ~nrow(.x)))
mahalanobis_distance <- function(x){
  x <- as.matrix(x)
  t(x) %*% x
}
mahalanobis_data <- mahalanobis_data %>% 
  mutate(d = map_dbl(data, ~mahalanobis_distance(.x)))
mahalanobis_data <- mahalanobis_data %>% 
  mutate(p_value = pchisq(d, df, lower.tail = FALSE))
mahalanobis_data %>% 
  filter(p_value < 0.05)
```
We can see that there are about 134 outliers who have p-values < 0.05.  With 1309 subjects, we would expect 1309 * 0.05 = 66.  We have around twice this amount, which may account for the heavy tails in the Q-Q plot. 

