---
title: "Residual Analysis"
author: "Hagen Atkeson, Mandy Woo, Allison Mariko Dayoan"
date: "5/21/2021"
output: html_document
---

```{r message = FALSE, warning = FALSE}
library(tidyverse)
library(magrittr)
library(nlme)
library(here)
library(mgcv)

aids <- read_csv(here::here("data/aids.csv"))
#aids <- read_csv("~/Documents/Stats-112-Final-Project/data/aids.csv")

aids %<>% 
  mutate(treatment = factor(treatment, levels = c(1, 2, 3, 4)),
         gender = factor(gender, levels = c("male", "female")))
```


When exploring the residual analysis for the model, we found that adding a quadratic term dramatically improved the semi-variogram.

```{r}
# create model

model_rand_q <- lme(log_cd4 ~ week + I(week^2) + treatment:week + gender:treatment + age:treatment,
                              data = aids,
                              random = ~ week | id,
                              method = "ML")

# get population residual
res_population <- residuals(model_rand_q, type = "response", level = 0)

# transformed residuals
Sigma_i <- extract.lme.cov(model_rand_q, aids)
L_i <- t(chol(Sigma_i)) 
res_transformed <- solve(L_i) %*% res_population

# histogram
tibble(r_star = res_transformed) %>% 
  ggplot(aes(x = r_star)) +
  geom_histogram(aes(y = stat(density)), bins = 14, color = "black", fill = "gray") +
  geom_function(fun = dnorm, color = "blue") +
  labs(x = "Residuals", y = "Density")

# QQPlot
tibble(r_star = res_transformed) %>% 
  ggplot(aes(sample = r_star)) +
  geom_qq_line(color = "blue") +
  geom_qq(shape = 1) +
  labs(x = "Quantiles of Standard Normal", y = "Quantiles of Transformed Residuals")

```
The QQPlot shows us that the transformed residuals have heavier tails than what would be expected for data from the Normal distribution, and that perhaps the assumption of normality is invalid.

```{r}
# scatter plot of predicted value vs. residuals

mu_hat <- fitted(model_rand_q, level = 0)
mu_hat_transformed <- solve(L_i) %*% mu_hat

tibble(x = mu_hat_transformed, y = res_transformed) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Transformed Predicted Value", y = "Transformed Residual")

# scatter plot of time vs. residuals

time <- aids$week
time_transformed <- solve(L_i) %*% time

tibble(x = time_transformed, y = res_transformed) %>% 
  ggplot(aes(x = x, y = y)) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE) +
  labs(x = "Transformed Time", y = "Transformed Residual")
```
Both scatterplots show that the mean of the transformed residuals stays approximately zero in regard to the predicted value and over time. This indicates that the assumed model of variance induced by the linear mixed-effe is adequate.

```{r}
# Semi-variogram
Variogram(model_rand_q,
          data = aids,
          form = ~ week | id,
          resType = "normalized") %>% 
  as_tibble() %>% 
  ggplot(aes(x = dist, y = variog)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_point(shape = 1) +
  geom_smooth(method = "loess", se = FALSE, span = 0.3)
```

```{r}
# Mahalanobis Distance

```

